\chapter{First Contact: Registers, Memory, and the Call}
\label{ch:asm1}

This chapter begins the computer architecture track. By its end you will have a
running \riscv assembly program on both QEMU and, if you have one, the physical
board---a program that clears a console, prints text, and makes a sound. More
importantly, you will understand the three ideas on which everything else rests:
what a register is, how a value lives in memory, and what happens when you call a
function. We build these ideas on the smallest possible game, the ASCII version
of Pong, exactly as the platform's own tutorial does \citep{prg32tutorialascii}.

\section{The machine you are programming}

A \riscv processor of the kind in your board is, from the programmer's point of
view, astonishingly simple. It has thirty-two registers, each holding a 32-bit
value, and it executes instructions that, with few exceptions, do one of three
things: an arithmetic or logical operation between registers, a transfer of a
value between a register and memory, or a change in which instruction runs next.
That is very nearly the whole machine. The richness of computing comes not from a
large instruction set but from combining a small one in long sequences.

\begin{prgconcept}
A register is a tiny, extremely fast storage location inside the processor that
holds one 32-bit value. The processor can only compute on values that are in
registers. To work on a value stored in memory, you must first \emph{load} it
into a register; to keep a result, you must \emph{store} it back to memory.
\end{prgconcept}

The thirty-two registers have conventional roles, fixed not by the hardware but
by an agreement called the \emph{calling convention} or \emph{ABI}. Following the
convention is what lets your assembly code and the framework's C code call each
other. Table~\ref{tab:registers} shows the registers you will actually use; the
full convention is in Appendix~\ref{app:api}.

\begin{center}
\begin{tabularx}{\linewidth}{@{}llX@{}}
\toprule
Register & Name & Role in this book \\
\midrule
\reg{a0}--\reg{a7} & arguments & values passed into and out of calls; \reg{a0} also holds the return value \\
\reg{t0}--\reg{t6} & temporaries & scratch values; a call may destroy them \\
\reg{s0}--\reg{s11} & saved & values a call promises to preserve \\
\reg{ra} & return address & where a \code{ret} jumps back to \\
\reg{sp} & stack pointer & top of the stack; kept 16-byte aligned \\
\reg{zero} & zero & always reads as 0 \\
\bottomrule
\end{tabularx}
\captionof{table}{The \riscv registers you will use, and their agreed roles
\citep{prg32abi}.}
\label{tab:registers}
\end{center}

\begin{prgconcept}
The \textbf{calling convention} is an agreement about which registers carry
arguments (\reg{a0}--\reg{a7}), which carry results (\reg{a0}), and which survive
a function call (\reg{s0}--\reg{s11}, but not \reg{t0}--\reg{t6}). \prg follows the
standard \riscv convention, which is exactly why an assembly game can call a C
helper such as \code{prg32\_gfx\_clear} \citep{prg32tutorial}.
\end{prgconcept}

\section{The smallest program that runs}

Recall the three-function contract from Chapter~\ref{ch:platform}. The smallest
legal \prg game provides \code{init}, \code{update}, and \code{draw} that each do
nothing but return. In \riscv assembly, ``return'' is the instruction \code{ret},
and ``do nothing else'' is simply not writing anything else. We name our game
\code{hello} and give every symbol that prefix.

\begin{lstlisting}[style=asm,caption={The smallest \prg assembly game: three functions that return immediately. Build it, confirm it runs, then add to it.}]
.option norelax
.section .text
.global hello_init
.global hello_update
.global hello_draw

hello_init:
    ret                 # return to the framework

hello_update:
    ret

hello_draw:
    ret
\end{lstlisting}

A few details deserve explanation. The directive \code{.option norelax} tells the
assembler not to apply an optimisation that can interfere with the way \prg
cartridges are linked; every example begins with it \citep{prg32tutorialgraphic}.
The \code{.section .text} directive says that what follows is executable code, as
opposed to data. The \code{.global} directives make the three symbols visible to
the framework, which must find them by name. Without \code{.global}, your
functions would exist but the loader could not see them.

\begin{prgcheck}
Build this skeleton for QEMU and confirm it runs without crashing. You will see
no game yet---that is correct. The achievement is that the framework found your
three functions, called them, and survived. Appendix~\ref{app:env} has the exact
build commands for your operating system; Appendix~\ref{app:worked} shows this
skeleton wired into the firmware.
\end{prgcheck}

\section{Calling a C helper: the stack frame}

A game that only returns is not interesting for long. The first useful thing we
can do is print text, using the framework's console. The console is written in C,
so calling it means crossing the boundary between your assembly and the
framework's C---and that boundary has one rule a beginner must internalise.

When you execute \code{call}, the processor records, in the register \reg{ra},
the address it should return to. But the C function you are calling will itself
use \reg{ra} for its own calls, overwriting yours. If you do nothing to protect
your \reg{ra}, then when your function tries to \code{ret}, it will jump to the
wrong place and your program will misbehave or crash. The solution is to save
\reg{ra} on the \emph{stack} before the call and restore it afterwards.

\begin{prgconcept}
The \textbf{stack} is a region of memory that grows downward and is used to save
values across function calls. Before calling a C helper, an assembly function
lowers the stack pointer \reg{sp} to make room, stores \reg{ra} there, makes its
call, then loads \reg{ra} back and raises \reg{sp}. This save/restore pair is the
single most important habit in \riscv assembly.
\end{prgconcept}

Here is the pattern, which you will repeat in nearly every function:

\begin{lstlisting}[style=asm,caption={The stack-frame idiom. Save \reg{ra}, do work that may call C, restore \reg{ra}, return.}]
hello_init:
    addi sp, sp, -16        # make a 16-byte frame (keeps sp 16-aligned)
    sw   ra, 12(sp)         # save the return address into the frame
    call prg32_console_clear   # clear the text console (a C helper)
    la   a0, title          # put address of the string in a0 (1st argument)
    call prg32_console_write   # print it
    lw   ra, 12(sp)         # restore the return address
    addi sp, sp, 16         # tear down the frame
    ret                     # now ret goes to the right place

.section .rodata
title:
    .asciz "HELLO PRG32\n"   # a zero-terminated string
\end{lstlisting}

Read this slowly, because it contains five distinct ideas. The \code{addi sp, sp,
-16} subtracts sixteen from the stack pointer, reserving sixteen bytes; we use a
multiple of sixteen because the convention requires \reg{sp} to stay 16-byte
aligned across calls \citep{prg32abi}. The \code{sw ra, 12(sp)} stores the word in
\reg{ra} at an offset of twelve bytes above the new stack top. The \code{la a0,
title} loads the \emph{address} of our string into \reg{a0}, because the
convention puts the first argument in \reg{a0} and the console function expects a
pointer. The two \code{call} instructions invoke C helpers. Finally the load and
\code{addi} undo the save and the reservation, in the reverse order, before
\code{ret}.

\begin{prgwarn}
Forgetting \code{sw ra} before a \code{call}, or \code{lw ra} after it, is the
classic first bug. The symptom is a function that ``returns to the wrong place'':
the program jumps somewhere unexpected, often crashing. If your assembly behaves
strangely around a call, check the \reg{ra} save/restore pair first
\citep{prg32tutorialascii}.
\end{prgwarn}

\begin{prgaside}
Why does the stack grow \emph{downward}, toward smaller addresses? It is a
convention almost as old as the stored-program computer, and it lets the stack
and the program's data grow toward each other from opposite ends of memory,
making the most of a fixed address space. Subtracting from \reg{sp} to allocate,
and adding to free, follows directly from that choice.
\end{prgaside}

\section{Values in memory: the \code{.data} section}

A game must remember things between frames---where the ball is, what the score
is. In assembly, persistent values live in the \code{.data} section, each given a
label and an initial value. This is the assembly equivalent of a global variable.
For ASCII Pong, the platform's example declares its state exactly this way
\citep{prg32repo}:

\begin{lstlisting}[style=asm,caption={Game state in \code{.data}. Each label names a memory location holding one 32-bit word.}]
.section .data
hello_x:      .word 20      # object x position, in text columns
hello_score:  .word 0       # a simple counter
\end{lstlisting}

To use such a value you do two things, and the distinction between them is the
crux of understanding memory. First you load the \emph{address} of the label into
a register with \code{la}. Then you load the \emph{value} at that address with
\code{lw}, or store a new value with \code{sw}.

\begin{lstlisting}[style=asm,caption={The load--modify--store cycle: read a value from memory, change it in a register, write it back.}]
    la t0, hello_x         # t0 = ADDRESS of hello_x
    lw t1, 0(t0)           # t1 = VALUE stored at that address
    addi t1, t1, 1         # change the value (in the register)
    sw t1, 0(t0)           # write the new value back to memory
\end{lstlisting}

\begin{prgconcept}
A label such as \code{hello\_x} names a \emph{location} in memory. \code{la t0,
hello\_x} puts that location's address in \reg{t0}; \code{lw t1, 0(t0)} reads the
\emph{contents} of that location into \reg{t1}. The processor computes only on
contents, in registers, then writes results back. This load--modify--store cycle
is how every variable in every program ultimately works.
\end{prgconcept}

\begin{prgwarn}
Keep the two clearly apart in your mind: \reg{t0} holds an \emph{address}, \reg{t1}
holds a \emph{value}. Confusing them---using an address where a value belongs, or
vice versa---produces wrong numbers and is one of the most common beginner errors.
A useful habit, used throughout the platform's examples, is to comment each
register with whether it holds an address or a value.
\end{prgwarn}

\section{Putting it together: a console game that ticks}

We now have every piece needed for a minimal but genuine game: an \code{init}
that prints a title, an \code{update} that advances a counter each frame, and a
\code{draw} that returns (we will draw properly in the next chapter). This is
deliberately close to the platform's own ASCII Pong skeleton
\citep{prg32tutorialascii}.

\begin{lstlisting}[style=asm,caption={A complete, minimal ASCII game. It runs identically on QEMU and the ESP32\nobreakdash-C6.}]
.option norelax
.section .data
hello_score:  .word 0

.section .rodata
title:  .asciz "HELLO PRG32\n"

.section .text
.global hello_init
.global hello_update
.global hello_draw

hello_init:
    addi sp, sp, -16
    sw   ra, 12(sp)
    call prg32_console_clear
    la   a0, title
    call prg32_console_write
    li   a0, 440            # 440 Hz: a short, pleasant beep on start
    li   a1, 60             # 60 ms duration
    call prg32_audio_beep   # silent on QEMU, audible on the board
    lw   ra, 12(sp)
    addi sp, sp, 16
    ret

hello_update:
    la   t0, hello_score
    lw   t1, 0(t0)
    addi t1, t1, 1          # count one tick per frame
    sw   t1, 0(t0)
    ret

hello_draw:
    ret
\end{lstlisting}

\begin{prgtargets}
\textbf{Both targets.} On QEMU you will see the title on the serial monitor; the
beep is accepted but silent because the emulator does not model the buzzer. On the
ESP32\nobreakdash-C6 you will additionally hear the beep through the buzzer. The
\emph{source is identical}; only the build directory and configuration profile
differ, as shown in Chapter~\ref{ch:platform}.
\end{prgtargets}

\begin{prgtry}
Change the beep frequency from \code{440} to \code{880} and rebuild. On the board,
the tone rises by an octave. Then change the increment in \code{hello\_update}
from \code{1} to \code{2} and predict, before running, what the counter does.
Verify your prediction. This ``change one thing, predict, verify'' loop is the
method of the entire book \citep{prg32repo}.
\end{prgtry}

\begin{prgcheck}
You should now be able to build and run an assembly program on at least QEMU;
explain why \reg{ra} must be saved before a \code{call}; explain the difference
between \code{la} and \code{lw}; and describe what the \code{.data}, \code{.text},
and \code{.rodata} sections are for. If any of these is shaky, re-read the
relevant section before Chapter~\ref{ch:asm2}; everything ahead builds on them.
\end{prgcheck}
