\chapter{State and Input: Branches, Bitmasks, and Boundaries}
\label{ch:asm2}

In Chapter~\ref{ch:asm1} a counter advanced on its own. A game, though, responds
to a player. This chapter teaches the assembly you need to read the controller
and to make decisions: the input bitmask, the bitwise \code{andi} test, and the
conditional branch. By the end you will move an object with the joystick and keep
it on the screen---the core interactive loop of every game in this book. We follow
the structure of the platform's ASCII Pong update routine, which was written to be
read by first-year students \citep{prg32tutorialascii,prg32repo}.

\section{Decisions: the conditional branch}

So far our instructions ran in a straight line. To respond to the player, a
program must choose: if a button is pressed, move; otherwise, do not. In \riscv
this choice is a \emph{conditional branch}, an instruction that jumps to a label
only when a condition holds.

\begin{prgconcept}
A \textbf{conditional branch} compares registers and jumps to a label only if the
comparison is true; otherwise execution simply continues to the next instruction.
The branch is how a processor makes every decision, from ``is this button
pressed?'' to ``has the ball hit the wall?''
\end{prgconcept}

The branches you will use most are few. \code{beqz t, label} jumps if \reg{t} is
zero; \code{bltz t, label} jumps if \reg{t} is negative; \code{ble a, b, label}
jumps if \reg{a} $\le$ \reg{b}; and the unconditional \code{j label} always jumps.
A common idiom, which surprises beginners, is to branch on the \emph{opposite} of
what you want: to do something ``if the button is pressed,'' you test the button
bit and \code{beqz}---branch away---when it is \emph{not} pressed, letting the
movement code run only when it is. This reads backwards at first but becomes
natural quickly.

\begin{prgaside}
Local labels written as a bare number followed by a colon, like \code{1:}, and
referenced as \code{1f} (the next \code{1} ``forward'') or \code{1b} (``backward''),
are an assembler convenience for short jumps that need no memorable name. The
platform's examples use them heavily for the small skips inside an update routine
\citep{prg32repo}.
\end{prgaside}

\section{Reading the controller}

Recall from Chapter~\ref{ch:platform} that \prg returns all buttons at once as a
bitmask. In assembly you obtain it by calling \code{prg32\_input\_read}; the
result arrives, by convention, in \reg{a0}. The very first thing to do is copy it
somewhere safe, because the next \code{call} you make would overwrite \reg{a0}.

\begin{lstlisting}[style=asm,caption={Read the input bitmask and keep it. The copy to a temporary is essential, because later calls reuse \reg{a0}.}]
    call prg32_input_read    # button bitmask returned in a0
    mv   t0, a0              # keep our own copy in t0 before anything else
\end{lstlisting}

\begin{prgwarn}
Copy the input mask out of \reg{a0} immediately. \reg{a0} is the argument and
return register; the moment you call another function, its old contents are gone.
The platform's examples copy the mask into a temporary on the very next line for
exactly this reason \citep{prg32tutorialascii}.
\end{prgwarn}

To test a single button you isolate its bit with \code{andi}, the
``and-immediate'' instruction, using the bit value from
Table~\ref{tab:buttons}. The result is non-zero exactly when that button is
pressed.

\begin{lstlisting}[style=asm,caption={Testing one button. \code{andi} keeps only the chosen bit; \code{beqz} skips the body when that bit is clear.}]
    andi t3, t0, 1          # PRG32_BTN_LEFT is bit 0 (value 1)
    beqz t3, 1f             # if the LEFT bit is clear, skip the move
    addi t1, t1, -1         # LEFT is pressed: move the object left
1:
\end{lstlisting}

\begin{prgconcept}
\code{andi t3, t0, 1} computes the bitwise AND of the input mask with the
constant \code{1}, leaving in \reg{t3} the value of bit~0 alone. If LEFT is
pressed, \reg{t3} is non-zero; if not, it is zero. Testing a button is therefore
``mask the bit, then branch on whether the result is zero.''
\end{prgconcept}

\section{Moving an object with the joystick}

We now assemble a complete \code{update} that moves an object left and right with
the joystick. The state is a single word in \code{.data}, exactly as in
Chapter~\ref{ch:asm1}. The routine reads input, tests LEFT and RIGHT, and updates
the position. This mirrors the platform's ASCII Pong update almost line for line
\citep{prg32repo}.

\begin{lstlisting}[style=asm,caption={A full \code{update}: read input, move on LEFT/RIGHT, store the result. Compare with the platform's \code{pong\_ascii\_update}.}]
.section .data
game_x:  .word 20           # object position, in text columns

.section .text
game_update:
    addi sp, sp, -16
    sw   ra, 12(sp)

    call prg32_input_read
    mv   t0, a0             # t0 = button bitmask

    la   t1, game_x
    lw   t2, 0(t1)          # t2 = current x (a value)

    andi t3, t0, 1          # test LEFT (bit 0)
    beqz t3, 1f
    addi t2, t2, -1         # move left
1:  andi t3, t0, 2          # test RIGHT (bit 1)
    beqz t3, 2f
    addi t2, t2, 1          # move right
2:  sw   t2, 0(t1)          # store the new x

    lw   ra, 12(sp)
    addi sp, sp, 16
    ret
\end{lstlisting}

Trace it once by hand. Suppose \code{game\_x} holds \code{20} and the player
holds RIGHT. The input read returns a mask with bit~1 set. The LEFT test
(\code{andi ... 1}) yields zero, so \code{beqz} skips the left move. The RIGHT
test (\code{andi ... 2}) yields non-zero, so the right move runs and \reg{t2}
becomes \code{21}. The store writes \code{21} back. Next frame it becomes
\code{22}, and so on. The object slides right as long as RIGHT is held. This kind
of by-hand trace is the most valuable debugging skill you will build in this
course.

\section{Staying on the screen: the boundary clamp}

Left unchecked, our object will slide right off the edge of the screen and
vanish. The fix is a \emph{clamp}: after computing the new position, force it back
into the legal range. For the 40-column ASCII console the legal range is 0 to 39;
for the 320-pixel graphics viewport it will be 0 to a value that accounts for the
object's width. The logic is identical either way.

\begin{lstlisting}[style=asm,caption={Clamping x to the range [0, 39]. Negative values snap to 0; values past the edge snap to the maximum.}]
    bltz t2, 3f             # if x < 0, clamp to the left edge
    li   t3, 39             # maximum column
    ble  t2, t3, 4f         # if x <= 39, it is already in range
    mv   t2, t3             # otherwise clamp to the right edge
    j    4f
3:  li   t2, 0              # left clamp
4:  sw   t2, 0(t1)          # store the bounded x
\end{lstlisting}

\begin{prgconcept}
A \textbf{boundary clamp} keeps a value inside a legal range: if it went below the
minimum, set it to the minimum; if above the maximum, set it to the maximum;
otherwise leave it alone. Almost every off-screen bug in a beginner's game is a
missing or wrong clamp. When an object disappears, suspect the clamp first.
\end{prgconcept}

\begin{prgwarn}
The maximum is not the screen width; it is the screen width minus the object's
own size. A 64-pixel-wide paddle on a 320-pixel screen may go no further right
than $320-64 = 256$, or its right edge would leave the viewport. The platform's
graphics Pong clamps the paddle to exactly \code{256} for this reason
\citep{prg32repo}.
\end{prgwarn}

\begin{prgtry}
Add the clamp above to your \code{update}. Then deliberately break it: change the
maximum from \code{39} to \code{60}, rebuild, and hold RIGHT. Watch the object
leave the visible area. Restore the clamp, rebuild, and confirm it stays. Write
one sentence explaining which comparison protects the right edge. This
break-and-fix exercise is taken directly from the platform's labs
\citep{prg32tutorialgraphic}.
\end{prgtry}

\section{Debugging like a systems programmer}

When an assembly game misbehaves, the platform's guidance is explicit and worth
adopting as a discipline: do not guess, trace state \citep{prg32tutorial}. Ask, in
order, which register held the last correct value, which instruction changed it,
whether a \code{call} overwrote a temporary you expected to keep, and whether a
branch skipped the code that updates memory. Reduce the program to the smallest
version that still shows the bug. Nine times in ten the answer is one of three
things: a missing \reg{ra} save, an address-versus-value confusion, or a branch
that goes the wrong way.

\begin{prgcheck}
You should now be able to read the input bitmask into a register; test an
individual button with \code{andi} and \code{beqz}; move a value in \code{.data}
in response to the joystick; and clamp that value to a legal range. With these,
you have written the interactive core of a game. Chapter~\ref{ch:asm3} adds the
one missing ingredient: drawing it on the screen.
\end{prgcheck}
