\chapter{Drawing the Game: Graphics, Collision, and the Frame}
\label{ch:asm3}

You can now move an object with the joystick and keep it on the screen. This
chapter gives that object a body. We move from the text console to the
320-by-200 graphics viewport, draw rectangles for a paddle and a ball, detect
when they touch, and tie the three functions together into a complete, playable
graphical game. The viewport calls behave identically on QEMU and on the
ESP32\nobreakdash-C6, so you can develop everything on a laptop and confirm it on hardware
later \citep{prg32qemu,prg32tutorialgraphic}.

\section{From text columns to pixels}

The console measured position in text columns; the viewport measures it in
pixels. The mental model is otherwise unchanged: state lives in \code{.data},
\code{update} changes it, \code{draw} renders it. What changes is the drawing
call. Instead of \code{prg32\_console\_write}, we use \code{prg32\_gfx\_clear} to
erase the frame and \code{prg32\_gfx\_rect} to draw a filled rectangle.

Recall the rectangle call from Chapter~\ref{ch:platform}:
\code{prg32\_gfx\_rect(x, y, w, h, color)}. In assembly the five arguments go,
in order, into \reg{a0} through \reg{a4}, following the calling convention. The
colour in \reg{a4} is an RGB565 value; you may write it as a named constant in C,
but in assembly it is usually a plain number---\code{65535} for white,
\code{65504} for yellow, \code{0} for black \citep{prg32repo}.

\begin{prgconcept}
The five arguments of \code{prg32\_gfx\_rect} occupy \reg{a0}=x, \reg{a1}=y,
\reg{a2}=width, \reg{a3}=height, \reg{a4}=colour. Getting an argument into the
wrong register draws the wrong rectangle---a vivid, immediate way to see what the
calling convention actually means.
\end{prgconcept}

\section{A blank frame, then one rectangle}

Following the platform's own first-game rhythm, we start with a \code{draw} that
clears the screen to black and does nothing else, confirm it runs, then add a
single rectangle \citep{prg32tutorialgraphic}.

\begin{lstlisting}[style=asm,caption={A \code{draw} that clears the frame. Build this first and confirm a black screen before drawing anything on it.}]
game_draw:
    addi sp, sp, -16
    sw   ra, 12(sp)
    li   a0, 0                 # RGB565 black
    call prg32_gfx_clear       # fill the 320x200 viewport
    lw   ra, 12(sp)
    addi sp, sp, 16
    ret
\end{lstlisting}

\begin{prgcheck}
Build this for QEMU and confirm the virtual screen shows solid black with no
crash. A black screen is success: the frame is being cleared every frame. If you
see garbage or the program hangs, recheck the \reg{ra} save/restore pair from
Chapter~\ref{ch:asm1}.
\end{prgcheck}

Now add a rectangle. We draw a white paddle near the bottom of the viewport, at
a position read from \code{.data}. The platform's graphics Pong places the paddle
at $y=188$, a 64-by-8 white bar---wide enough to be easy to see in the emulator
\citep{prg32repo}.

\begin{lstlisting}[style=asm,caption={Drawing the paddle from a stored position. The five arguments fill \reg{a0}--\reg{a4} in order.}]
    la   t0, paddle_x          # address of the paddle position
    lw   a0, 0(t0)             # a0 = x  (a value read from memory)
    li   a1, 188               # a1 = y  (near the bottom)
    li   a2, 64                # a2 = width
    li   a3, 8                 # a3 = height
    li   a4, 65535             # a4 = white (RGB565 0xffff)
    call prg32_gfx_rect        # draw the paddle
\end{lstlisting}

\begin{prgwarn}
The order of the loads matters because each \code{li} into \reg{a0}--\reg{a4}
sets one argument. If you load the width into \reg{a1} instead of \reg{a2}, you
draw a rectangle of the wrong shape in the wrong place. The platform's
break-and-fix lab asks you to do exactly this on purpose, observe the distorted
rectangle, then restore the order and explain the error
\citep{prg32tutorialgraphic}.
\end{prgwarn}

\section{The ball: position, velocity, and the bounce}

A paddle the player moves is half a game; the other half is a ball that moves on
its own and bounces. The ball needs four words of state: its $x$ and $y$ position
and its $x$ and $y$ velocity (which the examples call \code{dx} and \code{dy}).
Each frame, the velocity is added to the position; when the ball reaches an edge,
the velocity is negated so it travels back the other way. This is the heart of
the platform's graphics Pong update \citep{prg32repo}.

\begin{lstlisting}[style=asm,caption={Autonomous ball motion on the x axis with edge bounce. The same pattern, repeated for y, gives full 2-D motion.}]
.section .data
ball_x:  .word 156
ball_dx: .word 2               # +2 means moving right

.section .text
    la   t1, ball_x
    lw   t2, 0(t1)             # t2 = ball x
    la   t3, ball_dx
    lw   t4, 0(t3)             # t4 = ball dx (velocity)
    add  t2, t2, t4           # x = x + dx
    bltz t2, 1f                # hit the left edge?
    li   t5, 312               # 320 - ball width (8)
    ble  t2, t5, 2f            # still in range on the right?
    mv   t2, t5                # clamp to right edge
    neg  t4, t4                # reverse direction
    j    2f
1:  li   t2, 0                 # clamp to left edge
    neg  t4, t4                # reverse direction
2:  sw   t2, 0(t1)            # store new x
    sw   t4, 0(t3)            # store (possibly reversed) dx
\end{lstlisting}

\begin{prgconcept}
\textbf{Velocity} is just a number added to position each frame. A \emph{bounce}
is negating that number when the object reaches a boundary: \code{neg t4, t4}
turns $+2$ into $-2$, so the next frames move the ball back the way it came. With
this one idea---add velocity, negate at the edges---you can make anything bounce.
\end{prgconcept}

\begin{prgaside}
Notice that the bounce reuses the clamp from Chapter~\ref{ch:asm2}, with one
addition: where the clamp merely stopped the object at the edge, the bounce also
negates the velocity. The platform's examples keep the clamp so the ball never
visibly leaves the screen even for a single frame, then negate so it returns. Good
game feel often comes from such small combinations of simple rules.
\end{prgaside}

\section{Did they touch? Rectangle collision}

A ball that passes straight through the paddle is no fun. We need to know when
the ball's rectangle overlaps the paddle's rectangle. The framework provides
exactly this test: \code{prg32\_sprite\_hitbox(ax, ay, aw, ah, bx, by, bw, bh)}
takes the position and size of two rectangles and returns non-zero when they
overlap \citep{prg32tutorialgraphic}. In assembly the eight arguments fill
\reg{a0} through \reg{a7}.

\begin{lstlisting}[style=asm,caption={Testing ball-versus-paddle overlap. A non-zero result in \reg{a0} means they touched this frame.}]
    la   t0, ball_x
    lw   a0, 0(t0)            # ball x
    la   t0, ball_y
    lw   a1, 0(t0)            # ball y
    li   a2, 8                # ball width
    li   a3, 8                # ball height
    la   t0, paddle_x
    lw   a4, 0(t0)            # paddle x
    li   a5, 188              # paddle y
    li   a6, 64               # paddle width
    li   a7, 8                # paddle height
    call prg32_sprite_hitbox  # a0 = non-zero if overlapping
    beqz a0, 3f               # no overlap: skip the bounce
    la   t0, ball_dy          # overlap: make the ball go up
    li   t1, -1
    sw   t1, 0(t0)
3:
\end{lstlisting}

\begin{prgconcept}
Two rectangles overlap when they intersect on \emph{both} axes at once: the $x$
ranges overlap \emph{and} the $y$ ranges overlap. \code{prg32\_sprite\_hitbox}
performs this two-axis test for you and returns the answer as a single value, so
collision becomes one call followed by one branch.
\end{prgconcept}

\section{Assembling the complete game}

We now have every part: a paddle moved by input and clamped, a ball that moves
and bounces, and a collision test. Putting them in the three functions gives a
complete graphical game whose structure exactly matches the platform's
\code{pong\_graphics} example \citep{prg32repo}. The \code{init} prints a title
and beeps once; the \code{update} moves the paddle from input, advances and
bounces the ball, and applies collision; the \code{draw} clears the frame and
renders the paddle and ball.

\begin{lstlisting}[style=asm,caption={The \code{draw} of the complete game. Compare it line for line with \code{pong\_graphics\_draw} in the repository.}]
game_draw:
    addi sp, sp, -16
    sw   ra, 12(sp)
    li   a0, 0
    call prg32_gfx_clear       # black background

    la   t0, paddle_x          # draw the paddle
    lw   a0, 0(t0)
    li   a1, 188
    li   a2, 64
    li   a3, 8
    li   a4, 65535             # white
    call prg32_gfx_rect

    la   t0, ball_x            # draw the ball
    lw   a0, 0(t0)
    la   t0, ball_y
    lw   a1, 0(t0)
    li   a2, 8
    li   a3, 8
    li   a4, 65504             # yellow
    call prg32_gfx_rect

    lw   ra, 12(sp)
    addi sp, sp, 16
    ret
\end{lstlisting}

\begin{prgtargets}
\textbf{Both targets.} Develop and watch this on the QEMU virtual screen, where
the ball bounces and the paddle responds to injected input. Then build the
identical source for the ESP32\nobreakdash-C6 to play it with a real joystick and hear the
beep. Because QEMU defaults disable the physical GPIO buttons, the paddle may sit
still in the emulator unless input is injected; the ball still moves, so the
render path is fully verified on the desktop \citep{prg32qemu}.
\end{prgtargets}

\section{Step-by-step: how this game was built}

It is worth making explicit the pedagogical order in which a game like this is
constructed, because it is the order the platform recommends and the order in
which the example games in \code{examples/games} were themselves written
\citep{prg32tutorial,prg32tutorialgraphic}. You do not write the listing above in
one sitting; you grow it.

First, write \code{init}, \code{update}, and \code{draw} that only return, and
confirm the framework runs them. Second, add one \code{.data} variable for the
paddle position and draw the paddle at it. Third, read input in \code{update} and
move the paddle, then add its clamp. Fourth, add the ball's four state words and
its autonomous motion with bounces. Fifth, add the collision test. Only then,
sixth, add sound on the collision event. Each step is buildable and observable on
its own; if a step breaks, the bug is in the few lines you just added, not
hidden in a wall of untested code.

\begin{prgtry}
Build the game incrementally in exactly these six steps, rebuilding and running
after each. Keep a short log: after each step, write one line saying what you
added and what you saw. This log is not busywork---it is the written form of the
``change one thing, observe, record'' method, and it will make the difference
between a frustrating afternoon and a productive one
\citep{prg32repo}.
\end{prgtry}

\begin{prgcheck}
You can now clear the frame, draw rectangles from stored state, move the ball
with velocity and bounces, and detect collisions with one call. You have, in
other words, built a complete graphical game in \riscv assembly that runs on both
targets. Chapter~\ref{ch:asmcap} turns this into a capstone you extend and
package as a cartridge.
\end{prgcheck}
